---
title: "Progress Report"
author: "Tanisha Magikar"
date: "2026-4-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(rvest)
library(Lahman)
library(readr)
```

## Lab 4: Scraping Hall of Fame 2026 Data

### Goal

The goal of this report is to scrape the 2026 Baseball Hall of Fame voting data from Baseball Reference, clean the data into a usable format, save the cleaned data as a CSV file, and compare the results to the `HallOfFame` table in the `Lahman` package.

The final cleaned dataset should include:

- candidate name
- year
- votes
- percent of vote
- whether the candidate was inducted

### Scrape the Baseball Reference data

For this part, I used `rvest` to read the Hall of Fame voting page and collect the tables from the webpage.

```{r scrape-data}
hof_url <- "https://www.baseball-reference.com/awards/hof_2026.shtml"

hof_page <- read_html(hof_url)

hof_tables <- hof_page |>
  html_table(fill = TRUE)

length(hof_tables)
```

### Select the correct table

The page contains more than one table. The table I need is the BBWAA voting table, which is the first and largest table on the page.

```{r select-table}
table_summary <- tibble(
  table_number = seq_along(hof_tables),
  number_of_rows = map_int(hof_tables, nrow),
  number_of_columns = map_int(hof_tables, ncol)
)

table_summary
```

```{r raw-table}
hof_raw <- hof_tables[[1]]

glimpse(hof_raw)
head(hof_raw)
```

### Clean the column names

The scraped table does not immediately come in with perfect column names. The first row contains the useful column names, so I use that row as the actual header and then remove it from the data.

```{r clean-column-names}
new_column_names <- hof_raw[1, ] |>
  unlist(use.names = FALSE) |>
  as.character()

new_column_names <- ifelse(
  new_column_names == "" | is.na(new_column_names),
  paste0("extra_col_", seq_along(new_column_names)),
  new_column_names
)

new_column_names <- make.unique(new_column_names)

hof_clean_names <- hof_raw[-1, ]

names(hof_clean_names) <- new_column_names

glimpse(hof_clean_names)
head(hof_clean_names)
```

### Clean the data

The final dataset needs the candidate name, year, votes, percent of vote, and whether the candidate was inducted.

I also added a few extra columns that make the data easier to compare to Lahman: ballots, needed, votedBy, and category.

```{r clean-data}
year_of_vote <- 2026
total_ballots <- 425
votes_needed <- ceiling(total_ballots * 0.75)

hof_2026_clean <- hof_clean_names |>
  transmute(
    candidate = Name,
    yearID = year_of_vote,
    votes = parse_number(Votes),
    percent_vote = parse_number(`%vote`),
    ballots = total_ballots,
    needed = votes_needed,
    votedBy = "BBWAA",
    category = "Player"
  ) |>
  mutate(
    candidate = candidate |>
      str_remove("^X-") |>
      str_remove("\\*$") |>
      str_squish(),
    inducted = if_else(votes >= needed, "Y", "N")
  ) |>
  filter(
    !is.na(candidate),
    candidate != "",
    candidate != "Name",
    !is.na(votes),
    !is.na(percent_vote)
  ) |>
  arrange(desc(percent_vote))

hof_2026_clean
```

### Check who was inducted

A player needed at least 319 votes out of 425 ballots to be inducted.

```{r inducted-players}
hof_2026_clean |>
  filter(inducted == "Y")
```

### Save the cleaned dataset

```{r save-data}
final_hof_2026 <- hof_2026_clean |>
  select(candidate, yearID, votes, percent_vote, inducted, ballots, needed, votedBy, category)

write_csv(final_hof_2026, "TanishaMagikar_HallOfFame_2026.csv")

final_hof_2026
```

### Compare to Lahman

Next, I compared my scraped Baseball Reference data to the 2026 records in Lahman's `HallOfFame` table.

```{r lahman-2026}
lahman_2026 <- Lahman::HallOfFame |>
  filter(yearID == 2026)

lahman_2026
```

### Compare inducted counts

```{r compare-counts}
final_hof_2026 |>
  count(votedBy, inducted)

lahman_2026 |>
  count(votedBy, inducted)
```

### Create name keys for comparison

The Baseball Reference data uses candidate names, while Lahman uses `playerID`. To make comparison easier, I created a simple name key from the Lahman `People` table.

```{r name-keys}
make_name_key <- function(x) {
  x |>
    str_to_lower() |>
    str_replace_all("[^a-z ]", " ") |>
    str_squish()
}

lahman_names <- Lahman::People |>
  transmute(
    playerID,
    lahman_name = str_squish(paste(nameFirst, nameLast)),
    name_key = make_name_key(lahman_name)
  )

scraped_for_compare <- final_hof_2026 |>
  mutate(name_key = make_name_key(candidate))

lahman_for_compare <- lahman_2026 |>
  left_join(lahman_names, by = "playerID") |>
  mutate(
    name_key = make_name_key(lahman_name),
    lahman_inducted = inducted
  ) |>
  select(
    playerID,
    lahman_name,
    name_key,
    yearID,
    votedBy,
    ballots,
    needed,
    lahman_votes = votes,
    lahman_inducted,
    category,
    needed_note
  )
```

### Full comparison

```{r full-comparison}
comparison_2026 <- full_join(
  scraped_for_compare,
  lahman_for_compare,
  by = c("name_key", "yearID"),
  suffix = c("_scraped", "_lahman")
) |>
  mutate(
    comparison_name = coalesce(candidate, lahman_name),
    source = case_when(
      !is.na(candidate) & !is.na(lahman_name) ~ "appears in both",
      !is.na(candidate) & is.na(lahman_name) ~ "Baseball Reference only",
      is.na(candidate) & !is.na(lahman_name) ~ "Lahman only",
      TRUE ~ "check"
    )
  ) |>
  select(
    comparison_name,
    yearID,
    source,
    votes,
    percent_vote,
    inducted,
    lahman_votes,
    lahman_inducted,
    votedBy_lahman,
    needed_note
  ) |>
  arrange(source, desc(percent_vote), comparison_name)

comparison_2026
```

### Exporting the data set

```{r export-data}
final_hof_2026 <- hof_2026_clean |>
  select(candidate, yearID, votes, percent_vote, inducted, ballots, needed, votedBy, category)

readr::write_csv(final_hof_2026, file = "TanishaMagikar_HallOfFame_2026.csv")
getwd()
file.exists("TanishaMagikar_HallOfFame_2026.csv")
list.files(pattern = "\\.csv$")
```

### Notes

The Baseball Reference table needed cleaning because the first row contained the actual column names. After fixing the column names, I kept the variables needed for the final dataset: candidate name, year, votes, percent of vote, and inducted status.

I calculated `inducted` by checking whether the candidate had at least 319 votes. This was the number needed because there were 425 total ballots and a player needed at least 75 percent of the vote.

The Baseball Reference data and Lahman data do not match row-for-row. Baseball Reference shows the BBWAA ballot for 2026, while the Lahman `HallOfFame` table may include other voting groups, such as the Contemporary Baseball Era committee. Because of that, differences between the two sources are expected.
