---
title: "Progress Report"
author: "Ryan Horsey"
date: "2026-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Progress Report: Ryan Horsey

## Scrape the Data

I used the `rvest` package to read the 2026 Hall of Fame voting table from Baseball Reference.
The raw table has no proper column names. They are stored in the first row of data, so I
extract them and overwrite the column names, then remove that first row.

```{r}
library(rvest)

url <- "https://www.baseball-reference.com/awards/hof_2026.shtml"
html <- read_html(url)
tables <- html_table(html)

data <- tables[[1]]

# The true column names are stored in row 1
actual_col_names <- data[1, ]
colnames(data) <- actual_col_names
data <- data[-1, ]
```

## Clean the Data

### Question 1 — Which table?

`tables[[1]]` is the BBWAA ballot table for 2026, which is what we need.
`tables[[2]]` would be the Veterans/Contemporary Era committee (we can ignore that).

### Question 2 — Are variable names correct?

Yes, after the fix above the column names match the header row shown on the website
(e.g. `Name`, `Votes`, `%vote`, `YoB`, etc.).

### Question 3 — Which columns to keep?

The lab requires at minimum: candidate name, year, votes, percent of vote, and
whether the candidate was inducted. I'll compare to the `Lahman::HallOfFame` structure
to make sure I'm aligned:

```{r}
library(Lahman)
head(HallOfFame, 3)
```

I'll keep `Name`, `Votes`, and `%vote` from the scraped data, then construct
`yearID`, `inducted`, and `percent_vote` myself.

### Question 4 — Which variables should be numeric?

`Votes` comes in as character and needs to be converted. `%vote` has a `%` sign so I
use `parse_number()` from `readr` to strip it.

```{r}
library(readr)

data$Votes <- as.numeric(data$Votes)
data$percent_vote <- readr::parse_number(data$`%vote`)
```

### Question 5 — Who was inducted?

The induction threshold is 75% of ballots cast. Using `percent_vote >= 75` is the
cleanest approach because some players have suppressed vote totals (shown as NA),
and checking votes directly against a raw number would leave those rows with a
missing `inducted` value instead of `"N"`.

```{r}
data$inducted <- ifelse(data$percent_vote >= 75, "Y", "N")
```

### Remove "X-" prefix

Some player names are prefixed with `"X-"` on Baseball Reference to indicate they
were dropped from future ballots. We clean that off so names are readable.

```{r}
data$Name <- gsub("X-", "", data$Name)
```

### Final cleaned dataset

```{r}
library(dplyr)

hof_2026 <- data |>
  select(Name, Votes, percent_vote, inducted) |>
  mutate(
    yearID   = 2026,
    votedBy  = "BBWAA",
    category = "Player"
  )

head(hof_2026)
```

## Compare to Lahman

```{r}
lahman_2026 <- HallOfFame |>
  filter(yearID == 2026)

# How many rows in each source?
nrow(hof_2026)
nrow(lahman_2026)

# Inducted counts
hof_2026 |> count(inducted)
lahman_2026 |> count(votedBy, inducted)
```

The scraped BBWAA table has many more rows than the Lahman 2026 data. That's expected:
the Lahman `HallOfFame` table currently only contains the **Contemporary Baseball Era**
committee results for 2026 (8 rows), not the full BBWAA ballot. The two sources are
organized differently, so an exact row-for-row match is not expected.

The players inducted according to the BBWAA ballot (percent_vote >= 75):

```{r}
hof_2026 |> filter(inducted == "Y") |> select(Name, Votes, percent_vote)
```

## Export the Dataset

```{r}
readr::write_csv(hof_2026, file = "HallOfFame_2026.csv")
```
